---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r globaloptions, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  message=FALSE,
  warnings = FALSE,
  fig.align = "center",
  cache = TRUE)
```

## Loading and preprocessing the data
The following code allows to download data from the repository, but the chunk is
set to eval=FALSE, in order to avoid downloading each time the markdown is knitted.

``` {r download, eval=FALSE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile="data_activity.zip", method="curl")
unzip("data_activity.zip")
```

### 1. Load the data
Importing data into R.


``` {r readcsv}
activity <- read.csv("activity.csv")
```

### 2. Process/transform the data
Formatting dates variable as date.


``` {r process date}
activity$date <- as.Date(
  activity$date, 
  "%Y-%m-%d")
```



## What is mean total number of steps taken per day?
### 1. Calculate the total number of steps taken per day
We calculate the total steps taken each day, grouping activity by date and 
summing up the steps in each interval for each day.

``` {r daily_steps}
library(dplyr)
daily_steps <- activity %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps,
                              na.rm=TRUE
                              )
            )
```

### 2. Make a histogram of the total number of steps taken each day
Now we can generate and show the histogram.

``` {r plot daily_steps}
plot(x = daily_steps$date,
     y = daily_steps$total_steps,
     type = "h", 
     col = "steelblue",
     lwd = 4,
     lend = "square",
     main = "Steps taken each day",
     xlab = "Day",
     ylab = "Total steps",
     xaxt = "n"
)
dates_to_label <- seq(min(daily_steps$date), max(daily_steps$date+1), by = "1 month")
axis.Date(1, at = dates_to_label, format = "%b %Y") 
```

### 3. Calculate and report the mean and median of the total number of steps <br>taken per day
The mean and median of the total number of steps taken for day is calculated as follows.


``` {r avg_dsteps}
avg_dsteps <- activity %>%
  group_by(date) %>%
  summarise(mean_steps = mean(steps, 
                              na.rm=TRUE
                              ),
            median_steps = median(steps, 
                                  na.rm=TRUE
                                  )
            ) %>%
  ungroup()
names(avg_dsteps) <- c("Date", 
                       "Mean steps", 
                       "Median steps"
                       )
```

We now report the mean and median total number of steps taken for each day.


``` {r kable avg_dsteps}
knitr::kable(avg_dsteps)
```

## What is the average daily activity pattern?
### 1. Make a time series plot of the 5-minute interval and the average number of <br>steps taken, averaged across all days 
First, we create the dataset of the average number of steps taken for each 
5-minute interval, across all days.

``` {r daily_pattern}
 daily_pattern <- activity %>%
  group_by(interval) %>%
  summarise(average_steps = mean(steps,
                                na.rm=TRUE
                                )
            )
```

Plotting data.


``` {r plot daily_pattern}
plot(daily_pattern$interval, 
     daily_pattern$average_steps, 
     type = "l", 
     main = "Average steps per interval", 
     xlab="Interval", 
     ylab= "Average steps"
     )
```

### 2. Which 5-minute interval, on average across all the days in the dataset, <br> contains the maximum number of steps?
Calculating the interval which contains the maximum number of steps.


``` {r max interval}
daily_pattern$interval[[which.max(daily_pattern$average_steps)]]
```

## Imputing missing values
### 1. Calculate and report the total number of missing values in the dataset 
The following chunk calculates the total number of rows with missing values in activity
dataset.

``` {r rows_with_na}
rows_with_na <- activity[!complete.cases(activity), ]
nrow(rows_with_na)
```


### 2. Devise a strategy for filling in all of the missing values in the dataset
The strategy devised is to fill missing values of steps taken with the average
steps of the same interval across all days.
We now calculate the new column and put it in a new numeric vector, called 
filled_steps. Whenever the steps for a given interval are present, we just copy
such value into filled_steps corrispondent value.

``` {r filled_steps}
filled_steps <- numeric(length=nrow(activity))
for (i in 1:nrow(activity)) {
  if(is.na(activity$steps[[i]])) {
    empty_interval <- activity$interval[[i]]
    filled_steps[[i]] <- daily_pattern$average_steps[[which(daily_pattern$interval==empty_interval)]]
  }
  else filled_steps[[i]] <- activity$steps[[i]]
  }
```

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
Since we already created the new column, i.e. "filled_steps, in the previous 
paragraph, we now just have to create the new dataset as a copy of "activity" 
and then substitute the steps column with filled_steps.


``` {r activity_filled}
activity_filled <- activity
activity_filled$steps <- filled_steps
```

### 4. Make a histogram of the total number of steps taken each day and Calculate <br>and report the mean and median total number of steps taken per day. 
First, we calculate the total steps taken each day (as in the first assignment,
but starting from activity_filled instead of activity. 

``` {r daily_steps_filled}
library(dplyr)
daily_steps_filled <- activity_filled %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps))
```
Now we can generate and show the histogram.

``` {r plot daily_steps_filled}
plot(x = daily_steps_filled$date,
     y = daily_steps_filled$total_steps,
     type = "h", 
     col = "steelblue",
     lwd = 4,
     lend = "square",
     main = "Steps taken each day (filled NAs)",
     xlab = "Day",
     ylab = "Total steps",
     xaxt = "n"
)
dates_to_label <- seq(min(daily_steps_filled$date), max(daily_steps_filled$date+1), by = "1 month")
axis.Date(1, at = dates_to_label, format = "%b %Y") 
```

The mean and median of the total number of steps taken for day is calculated 
again as in the first assignment, again starting with activity_filled instead of
activity.


``` {r avg_dsteps_filled}
avg_dsteps_filled <- activity_filled %>%
  group_by(date) %>%
  summarise(mean_steps_f = mean(steps),
            median_steps_f = median(steps)
            ) %>%
  ungroup()
names(avg_dsteps_filled) <- c("Date", 
                              "Mean steps (filled NAs)", 
                              "Median steps (filled NAs)")
```

We now merge the two datasets of mean and median steps per day and report them, 
in order to be able to evaluate the differences.


``` {r dsteps_compare}
names(daily_steps) <- c("Date", "Total steps")
names(daily_steps_filled) <- c("Date", "Total steps (filled NAs)")
dsteps_compare_temp1 <- merge(daily_steps,
                            daily_steps_filled,
                            by="Date"
                            )
dsteps_compare_temp2 <- merge(avg_dsteps, 
                            avg_dsteps_filled, 
                            by="Date"
                            )
dsteps_compare <- merge(dsteps_compare_temp1,
                        dsteps_compare_temp2,
                            by="Date"
                            )
dsteps_compare <- dsteps_compare %>%
  relocate( "Mean steps (filled NAs)", 
            .after = "Mean steps"
            )
knitr::kable(dsteps_compare)
```

We can see that the effect of filling is that estimate values after filling NAs
are always valid values, without NAs or NaNs. The total number of steps per day is
also affected.

## Are there differences in activity patterns between weekdays and weekends?
### 1. Create a new factor variable in the dataset with two levels – “weekday” <br>and “weekend” 
We now add a column to activity_filled, indicating if the date is a weekday or 
weekend day.

``` {r weekdays}
library(lubridate)
activity_filled <- activity_filled %>%
  mutate(
    day_of_week_num = wday(date, week_start = 1),
    day_type = case_when(
      day_of_week_num %in% c(6, 7) ~ "Weekend",
      TRUE ~ "Weekday"
    ),
    day_type = factor(day_type, levels = c("Weekday", "Weekend"))
  ) %>%
  select(-day_of_week_num)
```
### 2. Make a panel plot containing a time series plot of the 5-minute interval<br> and the average number of steps taken, averaged across all weekday days or <br>weekend days

First, we create the dataset of the average number of steps taken for each 
5-minute interval, separating weekdays and weekend days.

``` {r daily_pattern_w}
 daily_pattern_w <- activity_filled %>%
  group_by(interval,day_type) %>%
  summarise(average_steps = mean(steps,
                                na.rm=TRUE
                                )
            )
```

Plotting data.


``` {r plot daily_pattern_w}
library(ggplot2)
plotweek <- ggplot(daily_pattern_w,
       aes(
         x=interval,
         y=average_steps
         )
       ) + 
  geom_line(colour = "lightblue")+
  facet_wrap(daily_pattern_w$day_type, 
                 nrow=2, 
                 ncol=1
             )  +
  labs(
    x = "Interval",
    y = "Number of steps"
  ) +
  theme(
    panel.background = element_rect(
      fill = NA, 
      colour = "black"
      ),
    strip.background = element_rect(fill = "#FFE4B5")
    )
print(plotweek)
```
